---
title: "Smoothing & Matrices"
author: "Sadia Boksh"
date: "28/11/2020"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dslabs)
library(dplyr)
library(rmarkdown)
library(tidyverse)
library(lubridate)
library(purrr) #map functions
library(pdftools)
```

```{r}
fn <- system.file("extdata", "RD-Mortality-Report_2015-18-180531.pdf", package="dslabs")
dat <- map_df(str_split(pdf_text(fn), "\n"), function(s){
	s <- str_trim(s)
	header_index <- str_which(s, "2015")[1]
	tmp <- str_split(s[header_index], "\\s+", simplify = TRUE)
	month <- tmp[1]
	header <- tmp[-1]
	tail_index  <- str_which(s, "Total")
	n <- str_count(s, "\\d+")
	out <- c(1:header_index, which(n==1), which(n>=28), tail_index:length(s))
	s[-out] %>%
		str_remove_all("[^\\d\\s]") %>%
		str_trim() %>%
		str_split_fixed("\\s+", n = 6) %>%
		.[,1:5] %>%
		as_tibble() %>% 
		setNames(c("day", header)) %>%
		mutate(month = month,
			day = as.numeric(day)) %>%
		gather(year, deaths, -c(day, month)) %>%
		mutate(deaths = as.numeric(deaths))
}) %>%
	mutate(month = recode(month, "JAN" = 1, "FEB" = 2, "MAR" = 3, "APR" = 4, "MAY" = 5, "JUN" = 6, 
                          "JUL" = 7, "AGO" = 8, "SEP" = 9, "OCT" = 10, "NOV" = 11, "DEC" = 12)) %>%
	mutate(date = make_date(year, month, day)) %>%
        dplyr::filter(date <= "2018-05-01")

head(dat)

dat = na.omit(dat)
any(is.na(dat))

str(dat)

```

Lets Use the loess() function to obtain a smooth estimate of the expected number of deaths as a function of date. 

```{r}
span = 60/nrow(dat)
fit <- loess(deaths ~ as.numeric(date), degree=1, span = span, data=dat)
```

We will plot this resulting smooth function using the span about two months long.
```{r}
dat %>% 
    mutate(smooth = fit$fitted, year = as.character(year(date))) %>%
    ggplot(aes(date, deaths)) +
    geom_point(size=3, alpha=0.5, color="grey")+
    geom_line(aes(date,smooth),lty=1, color="red", size=1)
```


Now we will plot smooth estimates against day of the year, all on the same plot, but with different colors for each year.

```{r}

?predict
  dat %>% 
    mutate(smooth = predict(fit), day = yday(date), year = as.character(year(date))) %>%
    ggplot(aes(day, smooth, col = year)) +
    geom_line(lwd = 2)
    dat %>% 
    mutate(smooth = predict(fit, as.numeric(date)), day = yday(date), year = as.character(year(date))) %>%
    ggplot(aes(day, smooth, col = year)) +
    geom_line(lwd = 2)
```
Suppose we want to predict 2s and 7s in the mnist_27 dataset with just the second covariate. Can we do this? On first inspection it appears the data does not have much predictive power.

In fact, if we fit a regular logistic regression the coefficient for x_2 is not significant!

This can be seen using this code:

```{r}
library(broom)
mnist_27$train %>% glm(y ~ x_2, family = "binomial", data = .) %>% tidy()

qplot(x_2, y, data = mnist_27$train)

head(as.numeric(mnist_27$train$y==7))
```

But by fitting a loess line to the data above and plotting the results, we can see there is predictive power and the conditional probability is non-linear.

u shaped curve means the line close to 1 has higher conditional probability of being 7 and x2 between .27 and .34 has higher conditional probability of getting 2

```{r}


fit <- loess(as.numeric(y==7) ~ x_2,degree=2,data=mnist_27$train)

mnist_27$train %>% 
    mutate(smooth = fit$fitted) %>%
    ggplot(aes(x_2, as.numeric(y==7))) +
    geom_point(size=3, alpha=0.5, color="grey")+
    geom_line(aes(x_2,smooth),lty=1, color="red", size=1)


```
# Matrices

```{r}
?matrix

      x <- matrix(rnorm(100*10), 100, 10)
      dim(x)
      nrow(x)
      ncol(x)
      head(x)
      ?sweep
      
      x <- x + seq(nrow(x))
      head(x)
      
      x <- -x
          x <- 1:ncol(x)
          rowMeans(x)
```

```{r}
class(mnist$train$images)
x <- mnist$train$images 
y <- mnist$train$labels
head(y)
class(x)
head(x,1)
new_x<- x[x>50 & x<205]
dim(x)
mean(rowMeans(x>50 & x<205))

rowSums(x>50 & x<205)/784/

y <- rowMeans(mnist$train$images>50 & mnist$train$images<205)
qplot(as.factor(mnist$train$labels), y, geom = "boxplot")

#boxplot(y,new_x)
```
