---
title: "Predicting Cicada Body Length using Bayesian Data Analysis"
author: "Sadia Boksh"
date: "13/11/2020"
output:
  github_document:
    html_preview: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library("rmarkdown")
library("rjags")
library("dplyr")
library("ggplot2")
library(grid)

```

# Data
Lets explore cicada data.

```{r Explore Data, echo=FALSE}
cicada_df = read.table(file="Cicada.txt", header=TRUE)
```
```{r}
head(cicada_df)
```

Structure of the data:    
```{r}
str(cicada_df)
```
### Plots
```{r Plots}

par(mfrow=c(2,2))
plot(BL ~ BW, data=cicada_df)
plot(BL ~ WW, data=cicada_df)
plot(BL ~ WL, data=cicada_df)

```
```{r }
cicada_df %>% ggplot(aes(x=G,y=BL,group=G))+
              geom_boxplot()+
              facet_grid(. ~ Species)
```

Now lets check if there are any missing values.
```{r}
any(is.na(cicada_df))

cicada_df$Species = cicada_df$Species+1
cicada_df$gender = cicada_df$G+1

data_jags = as.list(cicada_df)
```
Looks like none. Lets change the Species column values to start from 1 and create a new columns gender that will start from 1.

# Models

## Model 1
First start off with a simple linear model with all features as predictors and check its DIC.
```{r}
simple_string = " model {
    for (i in 1:length(BL)) {
        BL[i] ~ dnorm(mu[i], prec)
        mu[i] = int+
                  b[1]*BW[i] +
                  b[2]*WL[i]+
                  b[3]*WW[i]+
                  b[4]*gender[i]+
                  b[5] *Species[i] 
                  

    }
    int ~ dnorm(0.0, 1.0/1.0e4)
    
    for (j in 1:5) {
        b[j] ~ dnorm(0.0, 1/1e4) # has variance 1.0
    }

    prec ~ dgamma(1/2.0, 1*2/2.0)
    sig = sqrt(1 / prec)
} "


set.seed(92)

params = c("int", "b", "sig")

simple_mod = init_model(simple_string, data_jags,3)

sim_dic = dic.samples(simple_mod, n.iter = 1e3)
sim_dic

```
## Model 2

Second model is a hierarchical model  with random intercepts i.e each species has its own intercepts
```{r}
h_modx_string = " model {
    for (i in 1:length(BL)) {
        BL[i] ~ dnorm(mu[i], prec)
        mu[i] = a[Species[i]] +
                  b[1]*BW[i] +
                  b[2]*WL[i]+
                  b[3]*G[i] +
                  b_intx * WL[i] * G[i]


    }
    for (i in 1:max(Species)) {
       a[i] ~ dnorm(a0, prec_a)
    }

    a0 ~ dnorm(0.0, 1.0/1.0e4)
    prec_a ~ dgamma(1/2.0, 1*2.0/2.0)
    tau = sqrt( 1.0 / prec_a )

    b_intx ~ dnorm(0.0, 1/1e4)
    for (j in 1:3) {
        b[j] ~ dnorm(0.0, 1/1e4)
    }

    prec ~ dgamma(1/2.0, 1*2/2.0)
    sig = sqrt(1 / prec)
} "


set.seed(92)

params = c("a0","a", "b","b_intx", "sig", "tau")

h_modx = init_model(h_modx_string, data_jags,3)
hx_dic = dic.samples(h_modx, n.iter = 1e3)
hx_dic

```
## COnvergence Diagnostics
DIC is better for this model, lets check how well it can predict the cicada body lengths. Fist we need to simulate the model and check trace plots and convergence diagnostics.

### Trace Plots
```{r}
h_x1_sim = simulate_model(h_modx,params,5e5)
h_x1_csim =  combine_simu(h_x1_sim)
plot(h_x1_sim)
```
### Autocorrelation check
Now check the gelmen rubin diagnostics for convergence
```{r}
  gelman.diag(h_x1_sim)
  autocorr.plot(h_x1_sim, lag.max=1000)
  effectiveSize(h_x1_csim)
```
```{r}


pm_coef_so = colMeans(h_x1_csim)# posterior mean


X = cbind(data_jags$BW,data_jags$WL,data_jags$G,data_jags$WL*data_jags$G, data_jags$Species)
head(X)

X1=X[X[, 5] == 1,]
X2=X[X[, 5] == 2,]
X3=X[X[, 5]== 3,]

yhat1 = drop(pm_coef["a[1]"]+ X1[,c(1:4)] %*% pm_coef_so[c(5:8)])
resid1 = cicada_df$BL[which(cicada_df$Species == 1)] - yhat1
plot(resid1,xlab="index", ylab="resid1", main="Residual check for species 1 (tredecula)") # against data index
plot(yhat1, resid1,xlab="yhat1", ylab="resid1", main="Residuals vs preditions for species 1 (tredecula)")

head(yhat1)
head(cicada_df$BL[which(cicada_df$Species == 1)] )
qqnorm(resid1)

yhat2 = drop(pm_coef["a[2]"]+ X2[,c(1:4)] %*% pm_coef_so[c(5:8)])
resid2 = cicada_df$BL[which(cicada_df$Species == 2)] - yhat2
plot(resid2) # against data index
plot(yhat2, resid2)
qqnorm(resid2)

yhat3 = drop(pm_coef["a[3]"]+ X3[,c(1:4)] %*% pm_coef_so[c(5:8)])
resid3 = cicada_df$BL[which(cicada_df$Species == 3)] - yhat3
plot(resid3) # against data index
plot(yhat3, resid3)
qqnorm(resid3)




cicada_m = c(.20,24,1,24)
cicada_f = c(.20,24,0,0)
(n_sim = nrow(h_x1_csim))

head(h_mod_csim[,c(5:8)])
mu_pred_m1_so =  h_x1_csim[,"a[1]"] + h_x1_csim[,c(5:8)] %*% cicada_m
y_pred_m1_so =  rnorm(n=n_sim, mu_pred_m1_so, h_x1_csim[, "sig"])

mu_pred_f1_so =  h_x1_csim[,"a[1]"] + h_x1_csim[,c(5:8)] %*% cicada_f
y_pred_f1_so =  rnorm(n=n_sim, mu_pred_f1_so, h_x1_csim[, "sig"])

mu_pred_m2_so =  h_x1_csim[,"a[2]"] + h_x1_csim[,c(5:8)] %*% cicada_m
y_pred_m2_so =  rnorm(n=n_sim, mu_pred_m2_so, h_x1_csim[, "sig"])

mu_pred_f2_so =  h_x1_csim[,"a[2]"] + h_x1_csim[,c(5:8)] %*% cicada_f
y_pred_f2_so =  rnorm(n=n_sim, mu_pred_f2_so, h_x1_csim[, "sig"])

mu_pred_m3_so =  h_x1_csim[,"a[3]"] + h_x1_csim[,c(5:8)] %*% cicada_m
y_pred_m3_so =  rnorm(n=n_sim, mu_pred_m3_so, h_x1_csim[, "sig"])

mu_pred_f3_so =  h_x1_csim[,"a[3]"] + h_x1_csim[,c(5:8)] %*% cicada_f
y_pred_f3_so =  rnorm(n=n_sim, mu_pred_f3_so, h_x1_csim[, "sig"])

mean(y_pred_m3 > y_pred_f3)

par(mfrow=c(3,1))
plot(density(y_pred_m1_so),xlab="BL", ylab="density", main="Posterior distribution of BL (Species 1 (tredecula)) ", lty=2, xlim=c(10,35))
lines(density(y_pred_f1_so), col="black", lty=1)
legend("topleft", legend=c("tredecula M","tredecula F"),
       lty=c(2,1), col=rep(c("black"), each=2), bty="n")
plot(density(y_pred_m2_so), col="red" ,lty = 2,xlim=c(10,35), main="Posterior distribution of BL (Species 1 (tredecassini))")
lines(density(y_pred_f2_so), col="red" ,lty=1)
legend("topleft", legend=c("tredecassini M","tredecassini F"),
       lty=c(2,1), col=rep(c("red"), each=2), bty="n")
plot(density(y_pred_m3_so), col="blue", lty=2,xlim=c(10,35), main="Posterior distribution of BL (Species 1 (tredecim))")
lines(density(y_pred_f3_so), col="blue", lty=1)
legend("topleft", legend=c("tredecim M","tredecim F"),
       lty=c(2,1), col=rep(c("blue"), each=2), bty="n")
```


## MOdel 3

Third model is also a hierarchical model, but its intercept will depend on both species and gender. 
```{r}
h_mod_string = " model {
    for (i in 1:length(BL)) {
        BL[i] ~ dnorm(mu[i], prec)
        mu[i] = a[Species[i],gender[i]] +
                  b[1]*BW[i] +
                  b[2]*WL[i]+
                  b_intx*WL[i]*gender[i]
                  

    }
    for (i in 1:max(Species)) {
      for(j in 1:max(gender)){
       a[i,j] ~ dnorm(a0, prec_a)
      }
    }

    a0 ~ dnorm(0.0, 1.0/1.0e4)
    prec_a ~ dgamma(1/2.0, 1*2.0/2.0)
    tau = sqrt( 1.0 / prec_a )

    b_intx ~ dnorm(0.0, 1/1e4)
    b_intx_a ~ dnorm(0.0, 1/1e4)
    for (j in 1:2) {
        b[j] ~ dnorm(0.0, 1/1e4)
    }

    prec ~ dgamma(1/2.0, 1*2/2.0)
    sig = sqrt(1 / prec)
} "


set.seed(92)

params = c("a0","a", "b","b_intx", "sig", "tau")

h_mod = init_model(h_mod_string, data_jags,3)

h_dic =  dic.samples(h_mod, n.iter=1e3)
h_dic


```



```{r}
h_mod_sim = simulate_model(h_mod,params,1e5)
h_mod_csim =  combine_simu(h_mod_sim)
pm_coef = colMeans(h_mod_csim)# posterior mean


X = cbind(data_jags$BW,data_jags$WL,data_jags$gender,data_jags$WL*data_jags$gender, data_jags$Species)
head(X)

X1_f=X[X[, 5] == 1 & X[, 3] == 1,]
X1_m=X[X[, 5] == 1 & X[, 3] == 2,]
X2_f=X[X[, 5] == 2 & X[, 3] == 1,]
X2_m=X[X[, 5] == 2 & X[, 3] == 2,]
X3_f=X[X[, 5]== 3 & X[, 3] == 1,]
X3_m=X[X[, 5]== 3 & X[, 3] == 2,]

beta_coeffs =  pm_coef[c(8:10)]

species_1_female = X1_f[,c(1:2,4)] %*% beta_coeffs
species_1_male = X1_m[,c(1:2,4)] %*% beta_coeffs

yhat1_f = drop(pm_coef["a[1,1]"]+species_1_female)
resid1_f = cicada_df$BL[which(cicada_df$Species == 1 & cicada_df$gender == 1)] - yhat1_f
plot(resid1_f,xlab="index", ylab="resid1", main="Residual check for species 1 (tredecula)") # against data index
plot(yhat1_f, resid1_f,xlab="yhat1", ylab="resid1", main="Residuals vs preditions for species 1 (tredecula)")
qqnorm(resid1_f)

yhat1_m = drop(pm_coef["a[1,2]"]+species_1_male)
resid1_m = cicada_df$BL[which(cicada_df$Species == 1 & cicada_df$gender == 2)] - yhat1_m
plot(resid1_m,xlab="index", ylab="resid1", main="Residual check for species 1 (tredecula)") # against data index
plot(yhat1_m, resid1_m,xlab="yhat1", ylab="resid1", main="Residuals vs preditions for species 1 (tredecula)")
qqnorm(resid1_m)


cicada_m = c(.20,24,24)
cicada_f = c(.20,24,0)
(n_sim = nrow(h_mod_csim))

head(h_mod_csim[,c(5:7)])
mu_pred_m1 =  h_mod_csim[,"a[1,2]"] + h_mod_csim[,c(8:10)] %*% cicada_m
y_pred_m1 =  rnorm(n=n_sim, mu_pred_m1, h_mod_csim[, "sig"])

mu_pred_f1 =  h_mod_csim[,"a[1,1]"] + h_mod_csim[,c(8:10)] %*% cicada_f
y_pred_f1 =  rnorm(n=n_sim, mu_pred_f1, h_mod_csim[, "sig"])

mu_pred_m2 =  h_mod_csim[,"a[2,2]"] + h_mod_csim[,c(8:10)] %*% cicada_m
y_pred_m2 =  rnorm(n=n_sim, mu_pred_m2, h_mod_csim[, "sig"])

mu_pred_f2 =  h_mod_csim[,"a[2,1]"] + h_mod_csim[,c(8:10)] %*% cicada_f
y_pred_f2 =  rnorm(n=n_sim, mu_pred_f2, h_mod_csim[, "sig"])

mu_pred_m3 =  h_mod_csim[,"a[3,2]"] + h_mod_csim[,c(8:10)] %*% cicada_m
y_pred_m3 =  rnorm(n=n_sim, mu_pred_m3, h_mod_csim[, "sig"])

mu_pred_f3 =  h_mod_csim[,"a[3,1]"] + h_mod_csim[,c(8:10)] %*% cicada_f
y_pred_f3 =  rnorm(n=n_sim, mu_pred_f3, h_mod_csim[, "sig"])

mean(y_pred_m1 > y_pred_f1)


par(mfrow=c(3,1))
plot(density(y_pred_m1),xlab="BL", ylab="density", main="Posterior distribution of BL (Species 1 (tredecula)) ", lty=2, xlim=c(10,35))
lines(density(y_pred_f1), col="black", lty=1)
legend("topleft", legend=c("tredecula M","tredecula F"),
       lty=c(2,1), col=rep(c("black"), each=2), bty="n")
plot(density(y_pred_m2), col="red" ,lty = 2,xlim=c(10,35), main="Posterior distribution of BL (Species 1 (tredecassini))")
lines(density(y_pred_f2), col="red" ,lty=1)
legend("topleft", legend=c("tredecassini M","tredecassini F"),
       lty=c(2,1), col=rep(c("red"), each=2), bty="n")
plot(density(y_pred_m3), col="blue", lty=2,xlim=c(10,35), main="Posterior distribution of BL (Species 1 (tredecim))")
lines(density(y_pred_f3), col="blue", lty=1)
legend("topleft", legend=c("tredecim). M","tredecim). F"),
       lty=c(2,1), col=rep(c("blue"), each=2), bty="n")


```







## Helper Fucntions
```{r helper functions}
init_model = function(mod_string, data_jags, n_chains){
  mod = jags.model(textConnection(mod_string), data=data_jags, n.chains=n_chains)
  update(mod, 1e3) #burn in
  mod
}

simulate_model =  function(mod, params, n_iter){
  coda.samples(model=mod,variable.names=params,n.iter=n_iter)

}
combine_simu = function(mod_sim){
  as.mcmc(do.call(rbind, mod_sim))
}

diagnose_covergence = function(mod_sim){
  plot(h_mod2_sim[, c("a[1,1]","a[1,2]")])
    plot(var_sim)

  gelman.diag(h_mod2_sim)
  autocorr.diag(h_mod2_sim)
  autocorr.plot(h_modx2_sim)
  effectiveSize(h_modx2_csim)
}

```
