---
title: "Confounding"
author: "Sadia Boksh"
date: "03/01/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```

```{r}
library(dslabs)
data("research_funding_rates")
research_funding_rates
```

Construct a two-by-two table of gender (men/women) by award status (awarded/not) using the total numbers across all disciplines.

```{r}

df<- research_funding_rates %>% summarise(men_awarded = sum(awards_men), women_awarded=sum(awards_women), men_not_awarded =sum(applications_men - awards_men), women_not_awarded =  sum(applications_women - awards_women) ) 

two_by_two <-  data.frame(awarded= c('yes', 'no'), men=c(df$men_awarded, df$men_not_awarded), women=c(df$women_awarded, df$women_not_awarded))
two_by_two %>% summarise(percent_men=round(men/sum(men)*100,1), percent_women = round(women/sum(women)*100, 1))

two_by_two %>% select(-awarded) %>% chisq.test()
```

Null hypothesis:  there is no association between two variables
Alternative hypothesis: Assumes that there is an association between the two variables.

level of significance (p value): 0.05. If p_value < 0.05, reject null hypothesis, otherwise accept null hypothesis

As p-value is 0.05, we have to accept the null hypothesis that is the difference in funding is not significant.


There may be an association between gender and funding. But can we infer causation here? Is gender bias causing this observed difference? The response to the original paper claims that what we see here is similar to the UC Berkeley admissions example. Specifically they state that this "could be a prime example of Simpson’s paradox; if a higher percentage of women apply for grants in more competitive scientific disciplines, then an analysis across all disciplines could incorrectly show 'evidence' of gender inequality."

To settle this dispute, use this dataset with number of applications, awards, and success rate for each gender:



```{r}

dat <- research_funding_rates %>% 
      mutate(discipline = reorder(discipline, success_rates_total)) %>%
      rename(success_total = success_rates_total,
             success_men = success_rates_men,
             success_women = success_rates_women) %>%
      gather(key, value, -discipline) %>%
      separate(key, c("type", "gender")) %>%
      spread(type, value) %>%
      filter(gender != "total")
dat %>% ggplot(aes(x=discipline, y=success, col=gender, size=applications)) + geom_point()+theme(axis.text.x = element_text(angle = 45, vjust=0.5))
```

We can see that female applicants tends to apply for Medical Science and Social Science disciplines but the overall success rate is low in those two disciplines. Most of the difference in funding came from these two competitive disciplines.
